Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2022-11-05 paper/paper/SPIE/article.tex
[WIP]

@@

\subsubsection{Bilinguality Dataset}

@@

\subsubsection{Depression Dataset}

@@

\subsubsection{Learning Dataset}

@@

\subsubsection{Attention Dataset}

@@

\subsubsection{Osteoarthritis Dataset}
The \textit{Osteo} dataset\cite{tetreaultBrainConnectivityPredicts2016}
includes whiole-brain rs-fMRI scans of healthy, pain-free controls
(\code{nopain} condition), and individuals with knee osteoarthritis.
Osteoarthritic patients were treated for 2 weeks with either placebo
(\code{pain} condition) or duloxetine (\code{duloxetine} condition).
We use the three binary classification tasks \code{nopain v duloxetine},
\code{nopain v pain}, and \code{pain v duloxetine} for our analysis.

\subsubsection{Parkinson's Dataset}
The \textit{Park} dataset\cite{madhyasthaDynamicConnectivityRest2015} includes
subjects with non-demented Parkinson's Disease (PD) and
healthy controls performed a number of repetitions of the Attention Network
Task (ANT; \cite{fanActivationAttentionalNetworks2005}) during scans.
The classification task for this dataset is Parkinson's versus controls, i.e.
\code{park v ctrl}.

\subsection{Classification Task Subgroups}

For each dataset, we construct subgroups based either on existing labels in the
data, or for the \textit{Attention} dataset, from psychological measures available
from the data.

\begin{table}[h!]
\caption{
    ID = Identifier for paper. Subgroup = name of subgroup. Subjects = Number of subjects.
    Task = fMRI task. ANT = Attention Network Task \cite{fanActivationAttentionalNetworks2005}
}
\label{table:2}
\small
\centering
\begin{tabular}{ l l c c c }
\hline
\textbf{ID}  & \textbf{Subgroup}  & \textbf{Subjects}  & \textbf{Task} & \textbf{Scans per Subject} \\
\hline
AGING   &  older             & 28      & resting-state         & 1 \\
        &  younger           & 34      & resting-state         & 1 \\
        &                    &         &                       &   \\
BILING  &  bilingual          & 59      & resting-state        & 1 \\
        &  monolingual        & 33      & resting-state        & 1 \\
        &                    &         &                       &   \\
DEPRESS &  depression         & 51      & resting-state        & 1 \\
        &  control            & 21      & resting-state        & 1 \\
        &                    &         &                       &   \\
LEARN   &  task              & 24      & learn image sketches  & 16 \\
        &  rest              & 24      & resting-state         & 2 \\
        &                    &         &                       &   \\
OSTEO   &  duloxetine        & 19      & resting-state         & 1 \\
        &  pain              & 37      & resting-state         & 1 \\
        &  nopain            & 20      & resting-state         & 1 \\
        &                    &         &                       &   \\
PARK    &  Parkinson's       & 25      & ANT                   & 12 \\
        &  control           & 21      & ANT                   & 12 \\
        &                    &         &                       &   \\
PSYCH   &  vigilant           & 11     & resting-state        & 3 \\
        &  nonigilant         & 11     & resting-state        & 3 \\
        &  trait-attentive     & 11    & resting-state        & 3 \\
        &  trait-non-attentive & 11    & resting-state        & 3 \\
        &  task-attentive     & 11     & resting-state        & 3 \\
        &  task-nonattentive  & 11     & resting-state        & 3 \\
\hline
\end{tabular}
\end{table}

@@

Given empirically observed eigenvalues \(\Lambda\), the spectral rigidity, \(\Delta_3(L)\), is
calculated for any positive real value \(L < \max(\Lambda)\) as:
\[
\Delta_3(L) = \left \langle \min_{A,B} \frac{1}{L} \int_c^{c+L} \left(  \eta(\lambda) -A \lambda - B \right)^2 \right \rangle_c
\]
where \(\eta(\lambda)\) is the number of unfolded eigenvalues less than or equal to \(\lambda\),
\(\langle \cdot \rangle_c\) denotes the average with respect to all starting points \(c\), and where
\(A\) and \(B\) denote the slope and intercept, respectively, of the least squares fit of a straight
line to \(\eta(\lambda)\) on \([c, c+L]\) \cite{guhrRandommatrixTheoriesQuantum1998a}. The spectral
rigidty for a value \(L\) is thus the \textit{average nonlinearity}

@@

\section{Results}
\label{sect:results}

>>>>

\begin{table}[h!]
\caption{
    ID = Identifier for paper. Subgroup = name of subgroup. Subjects = Number of subjects.
    Task = fMRI task. ANT = Attention Network Task \cite{fanActivationAttentionalNetworks2005}
}
\label{table:2}
\small
\centering
\begin{tabular}{ l r c c c }
\hline
\textbf{ID}  & \textbf{Subgroup}  & \textbf{Subjects}  & \textbf{Task} & \textbf{Scans per Subject} \\
\hline
Aging   &  older             & 28      & resting-state        & 1 \\
        &  younger           & 34      & resting-state        & 1 \\
        &                    &         &                      &   \\
Bilingual  &  bilingual          & 59      & resting-state    & 1 \\
        &  monolingual        & 33      & resting-state       & 1 \\
        &                    &         &                      &   \\
Depress &  depression         & 51      & resting-state       & 1 \\
        &  control            & 21      & resting-state       & 1 \\
        &                    &         &                      &   \\
Learn   &  task              & 24      & learn image sketches & 16\\
        &  rest              & 24      & resting-state        & 2 \\
        &                    &         &                      &   \\
Osteo   &  duloxetine        & 19      & resting-state        & 1 \\
        &  pain              & 37      & resting-state        & 1 \\
        &  nopain            & 20      & resting-state        & 1 \\
        &                    &         &                      &   \\
Park    &  Parkinson's       & 25      & ANT                  & 12\\
        &  control           & 21      & ANT                  & 12\\
        &                    &         &                      &   \\
Attention&  vigilant           & 11     & resting-state       & 2 \\
        &  nonigilant         & 11     & resting-state        & 2 \\
        &  trait-attentive     & 11    & resting-state        & 2 \\
        &  trait-non-attentive & 11    & resting-state        & 2 \\
        &  task-attentive     & 11     & resting-state        & 2 \\
        &  task-nonattentive  & 11     & resting-state        & 2 \\
\hline
\end{tabular}
\end{table}

@@

% \subsubsection{Bilinguality Dataset}

@@

% \subsubsection{Depression Dataset}

@@

% \subsubsection{Learning Dataset}

@@

% \subsubsection{Osteoarthritis Dataset}
The \textit{Osteo} dataset\cite{tetreaultBrainConnectivityPredicts2016}
includes whiole-brain rs-fMRI scans of healthy, pain-free controls
(\code{nopain} condition), and individuals with knee osteoarthritis.
Osteoarthritic patients were treated for 2 weeks with either placebo
(\code{pain} condition) or duloxetine (\code{duloxetine} condition).
We use the three binary classification tasks \code{nopain v duloxetine},
\code{nopain v pain}, and \code{pain v duloxetine} for our analysis.

% \subsubsection{Parkinson's Dataset}
The \textit{Park} dataset\cite{madhyasthaDynamicConnectivityRest2015} includes
subjects with non-demented Parkinson's Disease (PD) and
healthy controls performed a number of repetitions of the Attention Network
Task (ANT; \cite{fanActivationAttentionalNetworks2005}) during scans.
The classification task for this dataset is Parkinson's versus controls, i.e.
\code{park v ctrl}.

% \subsubsection{Attention Dataset}

@@

@@

\subsection{fMRI Preprocessing}

We limited preprocessing to a simple pipeline of, in order, (1) brain
extraction, (2) slice timing correction, (3), motion correction, and (4),
registration to the MNI152
\cite{fonovUnbiasedAverageAgeappropriate2011,fonovUnbiasedNonlinearAverage2009}
2mm isotropic, asymmetric template version C made available through
\href{https://www.templateflow.org/}{TemplateFlow}
\cite{ciricTemplateFlowFAIRsharingMultiscale2021}.

We limited preprocessing methods to these steps sometimes because data required
for other typical preprocessing steps (e.g. field intensity maps, physiological
measurements), but also because we saved all pipelined intermediates to later
compare the effect of increasing degrees of pre-processing (see Section
\ref{sec:multiverse}), and it was important to limit the number of
preprocessing degrees to prevent excessive computational costs.

Brain extraction was performed first with BET
\cite{smithFastRobustAutomated2002}, and then with with
ANTs\cite{avantsReproducibleEvaluationANTs2011} to clean up residue left behind
by BET. BET results were visually inspected for each image in each dataset, and
BET fractional anisotropy values were modified to ensure brain extraction was
acceptable. Motion-correction was performed with FSL's MCFLIRT
\cite{jenkinsonImprovedOptimizationRobust2002}, and slicetime correction also
with FSL's \code{slicetimer}\cite{jenkinsonFSL2012} tool. Finally, functional
images were registered directly to the MNI152 template via
ANTs\cite{avantsReproducibleEvaluationANTs2011}.

\subsection{Feature Extraction}

For each preprocessing degree and fMRI image, features were extracted from all
\(N\) non-constant brain voxels, where \(N\) may differ for each image. After
this process, each scan yields an \(N \times T\) matrix \(\mathbf{M}\), where
\(T\) is the number of volumes acquired. All features summarize this matrix
of brain voxels \(\mathbf{M}\) in some manner.

\subsubsection{Baseline Features}

If RMT- or eigenvalue-based features fail to outperform simple reductions of
the fMRI data in predictive utility, it is difficult to justify the greater
computational and interpretational complexities of the former. These simple
statistical reductions help assess the relative value of RMT features.

Baseline reductions compute either a robust measure of location (mean, max,
min) or non-robust measure of location (median, \(95^{\text{th}}\) percentile,
\(5^{\text{th}}\) percentile), or a non-robust (standard deviation, range) or
robust (interquartile range, difference between \(95^{\text{th}}\) percentile
and \(5^{\text{th}}\) percentile) measure of scale summarizing across the voxel
dimension, yielding a feature of \(T\) dimensions. Exact details can be found
in the
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/enumerables.py#L51-L69}{paper
source code}.

\subsubsection{Baseline Features}
MAYBE MOVE THIS SECTION AFTER RMT AND EIGEN FEATURES?

fMRI global mean signal is

\subsubsection{Non-RMT Eigenfeatures}

Eigenfeatures summarize \(\mathbf{M}\) via some
transformation of the eigenvalues of the correlation matrix of \(\mathbf{M}\),
and thus yield a feature of  \(T - 1\) dimensions.

\subsubsection{RMT Features}
Given empirically observed eigenvalues \(\Lambda\), the spectral rigidity, \(\Delta_3(L)\), is
calculated for any positive real value \(L < \max(\Lambda)\) as:
\[
\Delta_3(L) = \left \langle \min_{A,B} \frac{1}{L} \int_c^{c+L} \left(  \eta(\lambda) -A \lambda - B \right)^2 \right \rangle_c
\]
where \(\eta(\lambda)\) is the number of unfolded eigenvalues less than or equal to \(\lambda\),
\(\langle \cdot \rangle_c\) denotes the average with respect to all starting points \(c\), and where
\(A\) and \(B\) denote the slope and intercept, respectively, of the least squares fit of a straight
line to \(\eta(\lambda)\) on \([c, c+L]\) \cite{guhrRandommatrixTheoriesQuantum1998a}. The spectral
rigidty for a value \(L\) is thus the \textit{average nonlinearity}

@@

\subsection{Multiverse Analysis}
\label{sec:multiverse}

\section{Results}
\label{sec:results}
